---
title: 'Illustrative applicationEstimating Optimal Decision Trees for Treatment Assignment:
  The Case of K > 2 Treatment Alternatives'
output:
  html_document:
    df_print: paged
---

This is a notebook to replicate the illustrative application of the paper "Optimal Decision Trees for Treatment Assignment: The Case of K > 2 Treatment Alternatives". 

```{r, echo=FALSE, warning=0, message=FALSE}
source('Multivalued code snippet.r')
library(quint)
library(partykit)
```

For this application, the BCRP data set (available in the quint r-package) is used. We start by loading this data
```{r, eval=TRUE,r, echo=FALSE, warning=FALSE}
dt <- na.omit(bcrp)
dt$wcht1=factor(dt$wcht1)
dt$nationality=factor(dt$nationality)
dt$marital=factor(dt$marital)
dt$cond=factor(dt$cond)
head(dt)
```


We start by estimating the misclassification costs using the AIPWE. To do so, we need an outcome- and propensity model. For the propensity model, we first use empirical proportions. 
```{r, eval=TRUE}
#Outcome model
set.seed(1)
model1 <- lm(I(physt3 - physt1) ~ cond*(1+ cesdt1 + physt1+ negsoct1 + uncomt1 + disopt1+ comorbid + age + wcht1+nationality+marital+trext), data=dt)

data1 <- dt
N=nrow(dt)
data1[,"cond"] <- as.factor(rep(1,N))
u1 <- predict.glm(model1, as.data.frame(data1))
data2 <- dt
data2[,"cond"] <- as.factor(rep(2,N))
u2 <- predict.glm(model1, as.data.frame(data2))
data3 <- dt
data3[,"cond"] <- as.factor(rep(3,N))
u3 <- predict.glm(model1, as.data.frame(data3))

#Propensity (empirical proportions)
set.seed(1)
PRaipwe <- table(dt$cond)/N  
cond=dt$cond
Y=dt$physt3-dt$physt1
#estimate contrast function
c1 <- (Y*(cond==1)/PRaipwe[1]) - (((cond==1) - PRaipwe[1]) / PRaipwe[1]*u1)
c2 <- (Y*(cond==2)/PRaipwe[2]) - (((cond==2) - PRaipwe[2]) / PRaipwe[2]*u2)
c3 <- (Y*(cond==3)/PRaipwe[3]) - (((cond==3) - PRaipwe[3]) / PRaipwe[3]*u3)
est.aipwe <- cbind(c1,c2,c3)

Zaipwe <- sapply(1:nrow(dt), function(b) which.max(est.aipwe[b,]))
dataCaipwe <- cbind(as.factor(Zaipwe),dt[,c(1,2,5:14)])

indiv.loss=t(sapply(1:nrow(dataCaipwe), function(b) max(est.aipwe[b,])-est.aipwe[b,]))
```

Next, we estimate the optimal treatment regime using the estimated misclassification costs, by means of the user defined splitting functions. We use five different seeds for pruning.
```{r, eval=TRUE}
## estimate optimal treatment regime
fit<-  rpart(as.numeric(Zaipwe) ~ cesdt1 + physt1+ negsoct1 + uncomt1 + disopt1+  
             comorbid + age + wcht1+nationality+marital+trext, 
             data = dataCaipwe, weights=1:nrow(dataCaipwe), method=multirisk, 
             parms=list(indiv.loss=t(sapply(1:nrow(dataCaipwe), function(b)   
             max(est.aipwe[b,])-est.aipwe[b,]))))


trees_aipwe=list(c(0))
for (i in 1:5){
  set.seed(111+i)
  xfit <- xpred.rpart(fit,xval=10)
  xerror <- colMeans(sapply(1:ncol(xfit), function(m) sapply(1:nrow(xfit),    
  function(k) indiv.loss[k,xfit[k,m]])))
  xerror
  cp1 <- fit$cptable[which.min(xerror),1]
  fitpr <- prune(fit, cp=cp1)
  trees_aipwe[[i]]=fitpr
  plot(as.party(trees_aipwe[[i]]))
}

```

Next, we use the theoretical proportions of 1/3 per treatment alternative as the propensity model and repeat the same procedure.
```{r, eval=TRUE}
#Propensity (empirical proportions)
set.seed(1)
PRaipwe <- c(1/3,1/3,1/3) 
cond=dt$cond
Y=dt$physt3-dt$physt1
#estimate contrast function
c1 <- (Y*(cond==1)/PRaipwe[1]) - (((cond==1) - PRaipwe[1]) / PRaipwe[1]*u1)
c2 <- (Y*(cond==2)/PRaipwe[2]) - (((cond==2) - PRaipwe[2]) / PRaipwe[2]*u2)
c3 <- (Y*(cond==3)/PRaipwe[3]) - (((cond==3) - PRaipwe[3]) / PRaipwe[3]*u3)
est.aipwe <- cbind(c1,c2,c3)

Zaipwe <- sapply(1:nrow(dt), function(b) which.max(est.aipwe[b,]))
dataCaipwe <- cbind(as.factor(Zaipwe),dt[,c(1,2,5:14)])

indiv.loss=t(sapply(1:nrow(dataCaipwe), function(b) max(est.aipwe[b,])-est.aipwe[b,]))

## estimate optimal treatment regime
fit<-  rpart(as.numeric(Zaipwe) ~ cesdt1 + physt1+ negsoct1 + uncomt1 + disopt1+  
             comorbid + age + wcht1+nationality+marital+trext, 
             data = dataCaipwe, weights=1:nrow(dataCaipwe), method=multirisk, 
             parms=list(indiv.loss=t(sapply(1:nrow(dataCaipwe), function(b)   
             max(est.aipwe[b,])-est.aipwe[b,]))))


trees_aipwe=list(c(0))
for (i in 1:5){
  set.seed(111+i)
  xfit <- xpred.rpart(fit,xval=10)
  xerror <- colMeans(sapply(1:ncol(xfit), function(m) sapply(1:nrow(xfit),    
  function(k) indiv.loss[k,xfit[k,m]])))
  xerror
  cp1 <- fit$cptable[which.min(xerror),1]
  fitpr <- prune(fit, cp=cp1)
  trees_aipwe[[i]]=fitpr
  plot(as.party(trees_aipwe[[i]]))
}

```

Finally, we use a Random Forest to estimate the misclassification costs. We use 3 different seeds for estimating the random forest-based contrasts. For estimating the optimal treatment regimes, we again use 5 seeds for pruning. 
```{r, eval=TRUE, warning=0, message=FALSE}
trees_rf=list(c(0))

library(ranger)
N=nrow(dt)
resp=dt$physt3-dt$physt1
k=1
for(i in 1:3){
  set.seed(1+i)
  RF= ranger(x=dt[,c(1,2,5:14)], y= resp ,always.split.variables="cond")
  data1 <- dt[,c(1,2,5:14)]
  data1[,"cond"] <- as.factor(rep(1,N))
  data2 <- dt[,c(1,2,5:14)]
  data2[,"cond"] <- as.factor(rep(2,N))
  data3 <- dt[,c(1,2,5:14)]
  data3[,"cond"] <- as.factor(rep(3,N))
  
  total=data.frame(rbind(data1,data2,data3))
  preds=predict(RF, total)
  est.rf = as.matrix(data.frame(u1=preds$predictions[1:224],       
                                u2=preds$predictions[225:448], 
                                u3=preds$predictions[449:672]))
  
  Zaipwe <- sapply(1:nrow(dt), function(b) which.max(est.rf[b,]))
  dataCaipwe <- cbind(as.factor(Zaipwe),dt[,c(1,2,5:14)])
  
  fit<-  rpart(as.numeric(Zaipwe) ~ cesdt1 + physt1+ negsoct1 + uncomt1 + disopt1+                 comorbid + age + wcht1+nationality+marital+trext, 
               data = dataCaipwe, weights=1:nrow(dataCaipwe), method=multirisk,
               parms=list(indiv.loss=t(sapply(1:nrow(dataCaipwe), 
                                              function(b) max(est.rf[b,])-est.rf[b,]))))
  indiv.loss=t(sapply(1:nrow(dataCaipwe), function(b) max(est.rf[b,])-est.rf[b,]))
  
  for (j in 1:5){
    set.seed(1+j)
    xfit <- xpred.rpart(fit, xval=10)
    xerror <- colMeans(sapply(1:ncol(xfit), function(m) sapply(1:nrow(xfit), FUN=       function(k) indiv.loss[k,xfit[k,m]])))
    xerror
    cp1 <- fit$cptable[which.min(xerror),1]
    fitpr <- prune(fit, cp=cp1)
    plot(as.party(fitpr))
    trees_rf[[k]]= fitpr
    k=k+1
  }
}

```
